### 2017-07-07 introduction.tex

The tension between serializable transactions, which are easy to
reason about but difficult to implement efficiently, and more
pragmatic variants that are driven by performance and availability
considerations, has motivated the development of weaker forms of
transaction isolation, beginning as early as 1976~\cite{gray1976}. The
ANSI SQL 92 standard defines three such weak isolation levels which
are now implemented in many relational and NoSQL databases. Not
surprisingly, weakly-isolated transactions have been found to
significantly outperform serializable transactions on benchmark
suites, both on single-node databases and multi-node replicated
stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to their
overwhelming adoption. A 2013 study~\cite{bailishotos} of 18 popular
ACID and ``NewSQL'' databases found that only three of them offer
serializability by default, and half, including Oracle 11g, do not
offer it at all.  A 2015 study~\cite{bailisferal} of a large corpus of
database applications finds no evidence that applications manifestly
change the default isolation level offered by the database. Taken
together, these studies make clear that weakly-isolated transactions
are quite prevalent in practice, whereas serializable transactions are
often eschewed.

@@

In this paper, we propose a program logic for weakly-isolated
transactions along with automated verification support that realizes
this goal.  We develop a set of syntax-directed compositional proof
rules that allow developers to build correctness proofs for
transactional programs in the presence of a weakly-isolated
concurrency control mechanism.  Realizing that the proof burden
imposed by these rules would likely discourage applications
programmers from using them, we present an inference procedure based
on a state transformer semantics that automatically verifies the
weakest isolation level for a transaction that ensures its consistency
requirements are maintained.  The key to inference is a novel formulation
of database state (represented as sets of tuples) as a monad, and in
which database computations are interpreted as transformers over these
sets.  This interpretation leads to an encoding of database
computations amenable for verification by off-the-shelf SMT solvers.

We have realized these ideas as an embedded DSL within OCaml, that
supports common relational SQL operations (updates, selects, inserts,
deletes, joins, etc.).  Experimental results on real-world benchmarks
demonstrate the feasibility and value of automated verification for
weakly isolated transactions, an important advance in improving the
safety of realistic database computations. The paper makes the
following contributions:
\vspace*{-6pt}
\begin{enumerate}
\item An embedded DSL in OCaml that treats SQL-based relational
  database operations as monadic computations over an abstract
  database state.
 \item An operational semantics and compositional proof system for
   this language capable of relating high-level application invariants
   to database state, parameterized by a weak isolation semantics that
   selectively exposes the visibility of these operations to other
   transactions.
 \item An inference algorithm capable of discovering the weakest
   isolation level that is sound with respect to a transaction's
   high-level consistency requirements.  The algorithm interprets
   database operations as state transformers expressed in a language
   amenable for translation into a decidable fragment of first-order
   logic, and is thus suitable for automated verification using
   off-the-shelf SMT solvers.
 \item A realistic implementation along with results on real-world
   database benchmarks that justify our approach and the utility of
   our proof methodology and inference procedure.
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
automated mechanism capable of verifying the safety of high-level
consistency conditions for weakly-isolated database transactions,
thereby allowing such transactions to enjoy the same rigorous
reasoning capabilities as their strongly-isolated (serializable)
counterparts.

>>>>

The tension between serializable transactions, which are easy to
reason about but difficult to implement efficiently, and more
pragmatic variants that are driven by performance and availability
considerations, has motivated the development of weaker forms of
transaction isolation, beginning as early as 1976~\cite{gray1976};
these instantiations permit a transaction to witness various effects
of newly committed, or even concurrently running, transactions while
it executes.  The ANSI SQL 92 standard defines three such weak
isolation levels which are now implemented in many relational and
NoSQL databases. Not surprisingly, weakly-isolated transactions have
been found to significantly outperform serializable transactions on
benchmark suites, both on single-node databases and multi-node
replicated stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to
their overwhelming adoption. A 2013 study~\cite{bailishotos} of 18
popular ACID and ``NewSQL'' databases found that only three of them
offer serializability by default, and half, including Oracle 11g, do
not offer it at all.  A 2015 study~\cite{bailisferal} of a large
corpus of database applications finds no evidence that applications
manifestly change the default isolation level offered by the
database. Taken together, these studies make clear that
weakly-isolated transactions are quite prevalent in practice, whereas
serializable transactions are often eschewed.

@@

In this paper, we propose a program logic for weakly-isolated
transactions along with automated verification support that realizes
this goal.  We develop a set of syntax-directed compositional proof
rules that allow developers to build correctness proofs for
transactional programs in the presence of a weakly-isolated
concurrency control mechanism.  Realizing that the proof burden
imposed by these rules would likely discourage applications
programmers from using them, we present an inference procedure based
on a state transformer semantics that automatically verifies the
weakest isolation level for a transaction that ensures its consistency
requirements are maintained.  The key to inference is a novel formulation
of database state (represented as sets of tuples) as a monad, and in
which database computations are interpreted as transformers over these
sets.  This interpretation leads to an encoding of database
computations amenable for verification by off-the-shelf SMT solvers.
The paper makes the following contributions:
%% We have realized these ideas as an embedded DSL within OCaml, that
%% supports common relational SQL operations (updates, selects, inserts,
%% deletes, joins, etc.).  Experimental results on real-world benchmarks
%% demonstrate the feasibility and value of automated verification for
%% weakly isolated transactions, an important advance in improving the
%% safety of realistic database computations.
\begin{enumerate}
\item We analyze properties of weak isolation in the context of an
  embedded DSL for OCaml that treats SQL-based relational database
  operations (e.g., inserts, selects, deletes, joins, updates, etc.)
  as monadic computations over an abstract database state.
 \item We develop an operational semantics and compositional proof
   system for this language capable of relating high-level application
   invariants to database state, parameterized by a weak isolation
   semantics that selectively exposes the visibility of these
   operations to other transactions.
 \item Exploiting the monadic definition of database state, we devise
   an inference algorithm capable of discovering the weakest isolation
   level that is sound with respect to a transaction's high-level
   consistency requirements.  The algorithm interprets database
   operations as state transformers expressed in a language amenable
   for translation into a decidable fragment of first-order logic, and
   is thus suitable for automated verification using off-the-shelf SMT
   solvers.
 \item We present details of a realistic implementation along with an
   evaluation study on real-world database benchmarks that justify our
   approach and the utility of our proof methodology and inference
   mechanism..
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
formalization of weakly-isolated transactions, along with an
expressive and compositional proof automation framework capable of
verifying the safety of high-level consistency conditions attached to
these transactions.  Collectively, these contributions allow
weakly-isolated transactions to enjoy the same rigorous reasoning
capabilities as their strongly-isolated (serializable) counterparts.

### 2017-07-07 motivation.tex

We present our ideas in the context of an embedded DSL in OCaml, which
offers a \C{DB} monad to define and compose database computations
given in terms of SQL operations.  Besides the usual \C{bind} and
\C{return} operators, the monad includes an \C{atomically} combinator
that executes a database computation as an atomic transaction and
returns the result.

@@

TPC-C defines multiple invariants, called \emph{consistency
conditions}, over the state of the application in the database. One
such consistency condition is the requirement that for a given order
\C{o}, the \emph{order-line-count} field (\C{o.o\_ol\_cnt}) should
reflect the number of order lines under the order; this is the number
of \C{Order\_line} records whose \C{ol\_o\_id} field is the same as
\C{o.o\_id}.  In a sequential execution, it is easy to see how this
condition is preserved.  A new \C{Order} record is added with its
\C{o\_id} distinct from existing order ids, and its \C{o\_ol\_cnt} is
set to be equal to the size of the \C{item\_reqs} set. The \C{foreach}
loop runs once for each \C{item\_req}, adding a new \C{Order\_line}
record for each requested item, with its \C{ol\_o\_id} field set to
\C{o\_id}. Thus, at the end of the loop, the number of \C{Order\_line}
records in the database, whose \C{ol\_o\_id} is equal to \C{o\_id} is
equal to the size of the \C{item\_req} set, which in turn is equal to
the \C{Order} record's \C{o\_ol\_cnt} field, thus preserving the
consistency condition.

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes.
Under weak isolation\footnote{Weak isolation does not violate
  atomicity as long as the witnessed effects are those of committed
  transactions}, however, interferences of various kinds are
permitted, leading to executions superficially similar to executions
permitted by concurrent (racy) programs~\cite{GHE15,HPQ+15}.
To illustrate, consider the behavior of the \C{new\_order} transaction
\begin{wrapfigure}{l}{.4\textwidth}
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits. Nonetheless, reading the same value for
  \C{d\_next\_o\_id} prompts both instances to add \C{Order} records
  with same ids, which in turn triggers a violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{wrapfigure}
when executed with a \emph{Read Committed} (RC) isolation level, the
default isolation level in 8 of the 18 databases studied
in~\cite{bailishotos}.  An RC transaction is isolated from \emph{dirty
  writes}, i.e., writes of uncommitted transactions, but is allowed to
witness the writes of concurrent transactions as soon as they are
committed. Thus, with two concurrent instances of the \C{new\_order}
transaction (call them $T_1$ and $T_2$), both concurrently placing new
orders for different customers under the same district (\C{d\_id}), RC
isolation allows the execution shown in Fig.~\ref{fig:new_order_exec}.

@@

The figure depicts an execution as a series of SQL 
operations. In the execution, the \C{new\_order} instance
$T_1$ (green) reads the \C{d\_next\_o\_id} field of the district
record for \C{d\_id}, but before it increments the field, another
\C{new\_order} instance ($T_2$) begins its execution and commits. Note
that $T_2$ reads the same \C{d\_next\_o\_id} value as $T_1$, and
inserts new \C{Order} and \C{Order\_line} records with their \C{o\_id}
and \C{ol\_o\_id} fields (resp.) equal to \C{d\_next\_o\_id}. $T_2$
also increments the \C{d\_next\_o\_id} field, which $T_1$ has already
acccessed. This is allowed because reads do not obtain a mutually
exclusive lock on most databases. After $T_2$'s commit, $T_1$ resumes
execution and adds new \C{Order} and \C{Order\_line} fields with the
same order id as $T_1$. Thus, by the end of the execution,
\C{Order\_line} records inserted by $T_1$ and $T_2$ all bear the same
order id. There are also two \C{Order} records with the same district
id (\C{d\_id}) and order id, none of whose \C{o\_ol\_cnt} reflects the
actual number of \C{Order\_line} records inserted with that order id.
This clearly violates TPC-C's consistency condition.

Notably, this example does not exhibit any of the anomalies that
characterize RC isolation~\cite{berenson}. For instance, there are no
\emph{lost writes}; both concurrent transactions' writes are present
in the final state of the database. Indeed, program analyses that aim
to determine appropriate isolation by checking for possible
manifestations of RC-induced anomalies would fail to identify grounds
for promoting the isolation level of \C{new\_order} to something
stronger.  Yet, if we take the semantics of the application into
account, it is quite clear that RC is not an appropriate isolation
level for \C{new\_order}.

While reasoning in terms of anomalies is cumbersome and inadequate,
reasoning about weak isolation in terms of low-level
traces~\cite{adyaphd,gotsmanconcur15} on memory read and write actions
can complicate high-level reasoning.  A possible alternative would
interleave weak isolation implementation details within the program,
yielding a (more-or-less) conventional concurrent program that can be
then subject to classical concurrent verification methods.
Considering the size and complexity of real-world transaction systems,
this strategy is unlikely to scale.

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
(SI), a stronger isolation level than RC. Snapshot isolation allows
transactions to be executed against a private snapshot of the
database, thus admitting concurrency, but it also requires that there
not be any write-write conflicts among concurrent transactions when
they commit. Write-write conflicts can be eliminated in various ways,
e.g., through conflict detection followed by a rollback, or through
exclusive locks, or a combination of both.  For instance, one possible
implementation of SI, close to the one used by
PostgreSQL~\cite{postgresiso}, executes a transaction against its
private snapshot of the database, but obtains exclusive locks on the
actual records in the database before performing writes. A write is
performed only if the record that is to be written has not already
been updated by a concurrent transaction. Conflicts are resolved by
abort and roll back.

@@

As this discussion indicates, implementations of SI on real databases
such as PostgresSQL are highly complicated, often running into
thousands of lines of code.  Nonetheless, the semantics of SI, in
terms of how it effects transitions on the database state, can be
captured in a fairly simple model. First, effects induced by one
transaction are not visible to another concurrently executing one
(call it $T$), during $T$'s execution. Thus, from $T$'s perspective,
the global state does not change during its execution.  More formally,
for every operation performed by $T$, the global state $T$ witness
before ($\stg$) and after ($\stg')$ executing the operation is the
same ($\stg'=\stg$).  After $T$ finishes execution, it commits its
changes to the actual database, which may have already incorporated
the effects of concurrent transactions. In executions where $T$
successfully commits, concurrent transactions are guaranteed to not be
in write-write conflict with $T$. Thus, if $\stg$ is the global state
that $T$ witnessed when it finished execution (the snapshot state),
and $\stg'$ is the state to which $T$ commits, then the difference
between $\stg$ and $\stg'$ should not result in a write-write conflict
with $T$. To concretize this notion, let the database state be a map
from transaction variables to values, and let $\stl$ denote a
transaction-local log that maps the variables being written to their
updated values.  The absence of write-write conflicts between $T$ and
the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  In other words, the
semantics of SI can be captured as an axiomatization over transitions
of the database state ($\Delta \longrightarrow \Delta'$) during the
lifetime an SI transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

\noindent This simple characterization of SI isolation allows us to
verify the consistency conditions associated with the \C{new\_order}
transaction.  First, since the database doesn't change ($\Delta' =
\Delta$) during the execution of the transaction's body, we can reason
about \C{new\_order} as though it executed in complete isolation until
its commit point, leading to a verification process similar to what
would been applied when reasoning about serializability.  When
\C{new\_order} finishes execution, but before it commits, the SI
axiomatization shown above requires us to consider global state
transitions $\stg \longrightarrow \stg'$ that do not include changes
to the records ($\stl$) written by \C{new\_order}, i.e.,
$\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.  The axiomatization
precludes any execution in which there may concurrent updates to
shared table fields (e.g., \C{d\_next\_o\_id} on the same \C{District}
table), but does not prohibit interferences that write to different
tables, or different records in the same table.  We need to reason
about the safety of such interferences with respect to the
transaction's consistency invariants.

@@

To do so, we observe that a relational database is a significantly
simpler abstraction than shared memory. There are no pointers, linked
data structures, or aliasing.  Although a database essentially
abstracts a mutable state, the state is affected through a
well-defined fixed number of interfaces (SQL statements), each tagged
with a logical formula describing what records are accessed and
updated.

This observation leads us away from thinking of database transactions
as concurrent imperative programs.  Instead, we see value in viewing
them as essentially functional computations that manage database state
monadically.  By doing so, we can formulate the semantics of database
operations as state transformers that relate the operation's pre- and
post-states.  A state transformer semantics can be defined
algorithmically, just like classical predicate transformer semantics
(e.g., strongest post-condition).  Here, a transformer interprets a
SQL statement in the set domain, modeling the database as a set of
tuples, and a SQL statement a function over this set.  Among other
things, one benefit of this approach is that low-level loops can now
be substituted with higher-order combinators that automatically lift
the state transformer of its higher-order argument (i.e., the loop
body) to the state transformer of the combined expression (i.e., the
loop).  We illustrate this intuition on a simple example.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a transaction
reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these updates. An update
statement filters the tuples that match the search criteria from $\Delta$
and computes the updated tuples that are to be
added to the local database. Thus, its state transformer (call it
$\T_U$) is the following function on sets\footnote{$\bind$ has higher
precedence than $\cup$.}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda s. 
      \itel{{\sf table}(s) = \C{Stock} \conj 
                s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}} 
           {\{\{s \with \C{s\_qty} = \C{k1}\}\}\\\hspace*{1.3in}}
           {\emptyset}
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$. 
The transformer ($\T_I(\Delta,\delta)$) for the subsequent \C{insert}
statement can be similarly constructed:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \{\C{ol\_o\_id}=\C{k2};\, 
        \C{ol\_d\_id}=\C{k3};\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id};\, 
        \C{ol\_qty}=\C{item\_req.ol\_qty}\}
\end{array}
\end{smathpar}
Observe that both transformers are of the form $T(\Delta,\delta) =
\stl \cup F(\stg)$, where $F$ is a function that returns the set of
records added by the transformer to the transaction-local database
($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for
$\T_U$ and $\T_I$ shown above. The state transformed of the loop body
in Fig.~\ref{fig:new_order_code} can be expressed as the following
composition of $\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stg) \cup \F_I(\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed to be:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stg) \cup \F_I(\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$,
leading to a completely mechanical inference procedure.  The advantage
of inferring such transformers is that we can now make use of a
semantics-preserving translation from the domain of sets equipped with
$\bind$ to first-order logic, allowing us to leverage SMT solvers for
automated proofs without having to infer potentially complex
thread-local invariants or intermediate assertions.
Sec.~\ref{sec:inference} describes this translation. In the exposition
thus far, we assumed $\Delta$ remains invariant, which is clearly not
the case when we admit concurrency.  Necessary concurrency extensions
of the state transformer semantics to deal with interference is also
covered in Sec.~\ref{sec:inference}.  The following two sections,
however, first focus on laying the theoretical foundations for
reasoning about weak isolation.

>>>>

We present our ideas in the context of an embedded DSL in OCaml that
we have developed, which offers a \C{DB} monad to define and compose
database computations given in terms of SQL operations.  Besides the
usual \C{bind} and \C{return} operators, the monad includes an
\C{atomically} combinator that executes a database computation as an
atomic transaction and returns the result.

@@

TPC-C defines multiple invariants, called \emph{consistency
  conditions}, over the state of the application in the database. One
such consistency condition is the requirement that for a given order
\C{o}, the \emph{order-line-count} field (\C{o.o\_ol\_cnt}) should
reflect the number of order lines under the order; this is the number
of \C{Order\_line} records whose \C{ol\_o\_id} field is the same as
\C{o.o\_id}.  In a sequential execution, it is easy to see how this
condition is preserved.  A new \C{Order} record is added with its
\C{o\_id} distinct from existing order ids, and its \C{o\_ol\_cnt} is
set to be equal to the size of the \C{item\_reqs} set. The \C{foreach}
loop runs once for each \C{item\_req}, adding a new \C{Order\_line}
record for each requested item, with its \C{ol\_o\_id} field set to
\C{o\_id}. Thus, at the end of the loop, the number of \C{Order\_line}
records in the database (i.e., the number of records whose
\C{ol\_o\_id} field is equal to \C{o\_id}) is guaranteed to be equal
to the size of the \C{item\_req} set, which in turn is equal to the
\C{Order} record's \C{o\_ol\_cnt} field; these constraints ensure that
the transaction's consistency condition is preserved.

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes,
essentially yielding serial behaviors.  Under weak
isolation\footnote{Weak isolation does not violate atomicity as long
  as the witnessed effects are those of committed transactions},
however, interferences of various kinds are permitted, leading to
executions superficially similar to executions permitted by concurrent
(racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
behavior of the \C{new\_order} transaction
\begin{wrapfigure}{l}{.4\textwidth}
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits. Nonetheless, reading the same value for
  \C{d\_next\_o\_id} prompts both instances to add \C{Order} records
  with same ids, which in turn triggers a violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{wrapfigure}
when executed with a \emph{Read Committed} (RC) isolation level, the
default isolation level in 8 of the 18 databases studied
in~\cite{bailishotos}.  An RC transaction is isolated from \emph{dirty
  writes}, i.e., writes of uncommitted transactions, but is allowed to
witness the writes of concurrent transactions as soon as they are
committed. Thus, with two concurrent instances of the \C{new\_order}
transaction (call them $T_1$ and $T_2$), both concurrently placing new
orders for different customers under the same district (\C{d\_id}), RC
isolation allows the execution shown in Fig.~\ref{fig:new_order_exec}.

@@

The figure depicts an execution as a series of SQL operations. In the
execution, the \C{new\_order} instance $T_1$ (green) reads the
\C{d\_next\_o\_id} field of the district record for \C{d\_id}, but
before it increments the field, another \C{new\_order} instance
($T_2$) begins its execution and commits. Note that $T_2$ reads the
same \C{d\_next\_o\_id} value as $T_1$, and inserts new \C{Order} and
\C{Order\_line} records with their \C{o\_id} and \C{ol\_o\_id} fields
(resp.) equal to \C{d\_next\_o\_id}. $T_2$ also increments the
\C{d\_next\_o\_id} field, which $T_1$ has already acccessed. This is
allowed because reads typically do not obtain a mutually exclusive
lock on most databases. After $T_2$'s commit, $T_1$ resumes execution
and adds new \C{Order} and \C{Order\_line} fields with the same order
id as $T_1$. Thus, by the end of the execution, \C{Order\_line}
records inserted by $T_1$ and $T_2$ all bear the same order id. There
are also two \C{Order} records with the same district id (\C{d\_id})
and order id, none of whose \C{o\_ol\_cnt} reflects the actual number
of \C{Order\_line} records inserted with that order id.  This clearly
violates TPC-C's consistency condition.

Notably, this example does not exhibit any of the anomalies that
characterize RC isolation~\cite{berenson}. For instance, there are no
\emph{lost writes} - both concurrent transactions' writes are present
in the final state of the database. Indeed, program analyses that aim
to determine appropriate isolation by checking for possible
manifestations of RC-induced anomalies would fail to identify grounds
for promoting the isolation level of \C{new\_order} to something
stronger.  Yet, if we take the semantics of the application into
account, it is quite clear that RC is not an appropriate isolation
level for \C{new\_order}.

While reasoning in terms of anomalies is cumbersome and inadequate,
reasoning about weak isolation in terms of
traces~\cite{adyaphd,gotsmanconcur15} on memory read and write actions
can complicate high-level reasoning.  A possible alternative would
interleave weak isolation implementation details within the program,
yielding a (more-or-less) conventional concurrent program that can be
then subject to classical concurrent verification methods.
Considering the size and complexity of real-world transaction systems,
this strategy is unlikely to scale.

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
(SI), a stronger isolation level than RC. Snapshot isolation allows
transactions to be executed against a private snapshot of the
database, thus admitting concurrency, but it also requires that there
not be any write-write conflicts (i.e,, both transactions modifying
the same record) among concurrent transactions when they
commit. Write-write conflicts can be eliminated in various ways, e.g.,
through conflict detection followed by a rollback, or through
exclusive locks, or a combination of both.  For instance, one possible
implementation of SI, close to the one used by
PostgreSQL~\cite{postgresiso}, executes a transaction against its
private snapshot of the database, but obtains exclusive locks on the
actual records in the database before performing writes. A write is
performed only if the record that is to be written has not already
been updated by a concurrent transaction. Conflicts are resolved by
abort and roll back.

@@

As this discussion indicates, implementations of SI on real databases
such as PostgresSQL are highly complicated, often running into
thousands of lines of code.  Nonetheless, the semantics of SI, in
terms of how it effects transitions on the database state, can be
captured in a fairly simple model. First, effects induced by one
transaction are not visible to another concurrently executing one
(call it $T$), during $T$'s execution. Thus, from $T$'s perspective,
the global state does not change during its execution.  More formally,
for every operation performed by $T$, the global state $T$ witness
before ($\stg$) and after ($\stg')$ executing the operation is the
same ($\stg'=\stg$).  After $T$ finishes execution, it commits its
changes to the actual database, which may have already incorporated
the effects of concurrent transactions. In executions where $T$
successfully commits, concurrent transactions are guaranteed to not be
in write-write conflict with $T$. Thus, if $\stg$ is the global state
that $T$ witnessed when it finished execution (the snapshot state),
and $\stg'$ is the state to which $T$ commits, then the difference
between $\stg$ and $\stg'$ should not result in a write-write conflict
with $T$. To concretize this notion, let the database state be a map
from transaction variables to values, and let $\stl$ denote a
transaction-local log that maps the variables being written to their
updated values.  The absence of write-write conflicts between $T$ and
the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  In other words, the
semantics of SI can be captured as an axiomatization over transitions
of the database state ($\Delta \longrightarrow \Delta'$) during a
transaction's ($T$) lifetime:
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

\noindent This simple characterization of SI isolation allows us to
verify the consistency conditions associated with the \C{new\_order}
transaction.  First, since the database does not change ($\Delta' =
\Delta$) during execution of the transaction's body, we can reason
about \C{new\_order} as though it executed in complete isolation until
its commit point, leading to a verification process similar to what
would been applied when reasoning sequentially.  When \C{new\_order}
finishes execution, however, but before it commits, the SI
axiomatization shown above requires us to consider global state
transitions $\stg \longrightarrow \stg'$ that do not include changes
to the records ($\stl$) written by \C{new\_order}, i.e.,
$\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.  The axiomatization
precludes any execution in which there are concurrent updates to
shared table fields (e.g., \C{d\_next\_o\_id} on the same \C{District}
table), but does not prohibit interferences that write to different
tables, or write to different records in the same table.  We need to
reason about the safety of such interferences with respect to
\C{new\_order}'s consistency invariants.

@@

We approach the verification problem by first observing that a
relational database is a significantly simpler abstraction than shared
memory. There are no pointers, linked data structures, or aliasing.
Although a database essentially abstracts a mutable state, this state
is affected through a well-defined fixed number of interfaces (SQL
statements), each tagged with a logical formula describing what
records are accessed and updated.

This observation leads us away from thinking of a collection of
database transactions as a simple variant of a concurrent imperative
program.  Instead, we see value in viewing them as essentially
functional computations that manage database state monadically,
mirroring the structure of our DSL.  By doing so, we can formulate the
semantics of database operations as state transformers that explicitly
relate an operation's pre- and post-states, defining the semantics of
the corresponding transformer algorithmically, just like classical
predicate transformer semantics (e.g., weakest pre-condition or
strongest post-condition).  In our case, a transformer interprets a
SQL statement in the set domain, modeling the database as a set of
tuples, and a SQL statement as a function over this set.  Among other
things, one benefit of this approach is that low-level loops can now
be substituted with higher-order combinators that automatically lift
the state transformer of its higher-order argument (i.e., the loop
body) to the state transformer of the combined expression (i.e., the
loop).  We illustrate this intuition on a simple example.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the tuples that match the search
criteria from $\Delta$ and computes the updated tuples that are to be
added to the local database. Thus, the state transformer for the update
statement (call it $\T_U$) is the following function on
sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda s. 
      \itel{{\sf table}(s) = \C{Stock} \conj 
                s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}} 
           {\{\{s \with \C{s\_qty} = \C{k1}\}\}\\\hspace*{1.3in}}
           {\emptyset}
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$. 
The transformer ($\T_I(\Delta,\delta)$) for the subsequent \C{insert}
statement can be similarly constructed:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \{\C{ol\_o\_id}=\C{k2};\, 
        \C{ol\_d\_id}=\C{k3};\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id};\, 
        \C{ol\_qty}=\C{item\_req.ol\_qty}\}
\end{array}
\end{smathpar}
Observe that both transformers are of the form $T(\Delta,\delta) =
\stl \cup F(\stg)$, where $F$ is a function that returns the set of
records added by the transformer to the transaction-local database
($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for
$\T_U$ and $\T_I$ shown above. The state transformation induced by the
loop body in Fig.~\ref{fig:new_order_code} can be expressed as the
following composition of $\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stg) \cup \F_I(\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed to be:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stg) \cup \F_I(\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
The advantage of inferring such transformers is that we can now make
use of a semantics-preserving translation from the domain of sets
equipped with $\bind$ to a decidable fragment of first-order logic,
allowing us to leverage SMT solvers for automated proofs without
having to infer potentially complex thread-local invariants or
intermediate assertions.  Sec.~\ref{sec:inference} describes this
translation. In the exposition thus far, we assumed $\Delta$ remains
invariant, which is clearly not the case when we admit concurrency.
Necessary concurrency extensions of the state transformer semantics to
deal with interference is also covered in Sec.~\ref{sec:inference}.
The following two sections, however, first focus on laying the
theoretical foundations for reasoning about weak isolation.
